PMV: Which formulation should you use the ISO 7730:2005 or ASHRAE 55:2020?

### 2023-09-01 Manuscript/src/sections/results.tex
paper: meeting with Stefano

\subsection{\ac{db2} overview}\label{subsec:comfort-db-overview}
While the \ac{db2} contains \var{entries_db_all} data points, only \var{entries_db_used} entries met the inclusion criteria listed in the Methodology section.
The distribution of the six input variables used to calculate the PMV indices is visualized in Figure~\ref{fig:dist_input_data}.
The 2.5th, 50th (median), and 97.5th percentiles are shown in blue.
% TODO:
%  remove the gridlines or marker them very short
%  the chart is a bit dense of information
\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/dist_input_data}
    \caption{Distribution of the input variables used to calculate the \ac{pmv} and \ac{pmv-ce} values.
    The text in blue is showing the 2.5th, 50th (median), and 97.5th percentiles.
    The data are shown using boxen plots (letter-value plots).
    They depict the median as the center-line and each successive level outward contains half of the remaining data.}
    \label{fig:dist_input_data}
\end{figure*}
The input data distributions are far from uniformity (equal density) across the Standards' applicability ranges.
We were not able to determine the accuracy of the \ac{pmv} and \ac{pmv-ce} formulations for inputs that were lower than the 2.5th and 97.5th percentiles, since we only had a limited number of data points beyond those values.
For example, only 2.5\% of the available dataset had a \ac{tdb} $\leq$ than \var{ta_95_perc_min}, this may consequently be an issue since both thermal comfort Standards' applicability limits extend to \qty{10}{\celsius}.
This can be explained by the fact that only a few buildings in the world operate at \ac{tdb} lower than this value.
This is because an indoor \ac{tdb} lower than \var{ta_95_perc_min} would be deemed to be unsatisfactory by a great number of occupants~\cite{iso7730}.
This lack of data in the \ac{db2} is particularly relevant for \ac{v}.
The \ac{pmv} and \ac{pmv-ce} only differ when the value of \ac{v} $\geq$ \qty{0.1}{\m\per\s}, however, the median value in the \ac{db2} was \qty{0.08}{\m\per\s} suggesting that most of the data points were collected in buildings with low air movement.
Consequently, we could not reliably test the accuracy of the \ac{pmv} formulations for values of \ac{v} $\geq$ \var{v_95_perc_max}.

Figure~\ref{fig:dist_other_data} shows the distribution of age, height, weight, and running mean outdoor temperature.
Less than half (\num{23300}) of the total entries have information about the participant's sex and are almost equally distributed among males \qty{52}{\percent} and females.
Ages are not normally distributed, the same is true for the running mean outdoor temperature.
About half of the participants (\qty{52}{\percent}) were aged between \num{20} and \num{35} years old, and only \qty{2.5}{\percent} were older than 60.
This limits our analysis to young adults and does not allow us to determine the accuracy of the \ac{pmv} and \ac{pmv-ce} models in predicting thermal sensation for children or older adults.
The \ac{db2} does not contain information about the health status of the participants.
Approximately \qty{56}{\percent} of the running mean outdoor temperature values where between \qtyrange{10}{25}{\celsius}.
% todo:
%  I merged men and women since not considering sex difference in the paper.
%  This figure can be merged with figure 2 because it is a visualization of the database.
%  To be more consisten in the paper, I would use the letter value plots for this and the figure in section 3.3.2
% remove the gridlines or marker them very short;
\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/dist_other_data}
    \caption{Distribution of age, height, weight, and running mean outdoor temperature.
    The data are grouped by sex.
    The text in blue is showing the 2.5th, 50th (median), and 97.5th percentiles.}
    \label{fig:dist_other_data}
\end{figure*}
The percentage of \ac{tpv} grouped by each \ac{tsv} is shown on the left side of Figure~\ref{fig:bar_plot_tp_by_ts}.
Only \var{entries_with_tp} entries had information about both \ac{tsv} and \ac{tpv}.
Thus, on the right, we show a bar chart depicting the distribution of all the \ac{tsv} votes, we analyzed in the paper.
The \ac{tsv} is unbalanced. 
Approximately \var{perc_tsv_neutral} of all the entries have a \ac{tsv} of `Neutral'.
While less than \var{perc_tsv_hot} of the total sample of participants reported to be either `Hot' or `Cold'.
% todo:
%  FT - I like my original colour more
%  SS - TSV chart the color here should not match the one on the thermal preference scale given that it is a different question. For example, a TSV=3 is red, like people that wants to be warmer, meaning that they are cold. Even with the inversion of the color palette, we cannot imply that TSV=3 mease wanted to be cooler. What about TSV=1, maybe people want no changer or cooler or warmer. Difficult to assume. Better to use two independet color schemes.
%  FT - then also Michael color scheme is wrong since we cannot assume that someone that want to be warm must be cold, they could be already slightly warm but at the same time they may want to be warmed.
\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/bar_plot_tp_by_ts}
    \caption{The left Figure shows the percentage of \ac{tpv} for each thermal sensation vote.
    The numbers on the right side of each bar show the number of points for each \ac{tsv}.
    The right figure shows the total number of data points grouped by \ac{tsv}.
    Each bin in the right figure has more data points than in the left figure since less data on \ac{tpv} were available.}
    \label{fig:bar_plot_tp_by_ts}
\end{figure*}

@@

\subsection{Comparison of PMV accuracy in predicting thermal sensation}\label{subsec:model-accuracy-comparison-in-predicting-thermal-sensation}
The \ac{pmv} was developed with the primary aim of predicting \ac{tsv}, consequently, we are here comparing the prediction accuracy of the two formulations.
% todo:
%  We can use Toby's or Michael but not both, even, though I like mine more
%  It is hard to notice that these are two distint graph because the legend is moving accross.
%  and have a thin grey line to separate the two group, another option would be to add a light grey background for the two PMV and show that they are distinct.
\begin{figure*}[htb!]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/bar_stacked_model_accuracy_0}
        \caption{All data}
     \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/bar_stacked_model_accuracy_0.2}
        \caption{Only data with \ac{v} $\geq$ \qty{0.2}{\m\per\s}}
     \end{subfigure}
    \caption{The calculated \ac{pmv} value is grouped by the \ac{tsv} value reported by the participants. 
    The number of data points in each bin of the two charts is the same and this allows us to compare the accuracy of both models.
    The overall prediction accuracy is shown above the chart while the model accuracy of predicting participants grouped by their \ac{tsv} is overlayed over each bar. 
    A random model, randomly selecting a thermal sensation vote would have an overall accuracy of \qty{14}{\percent}.}
    \label{fig:bar_stacked_model_accuracy}
\end{figure*}
Being \ac{tsv} the value reported by participants (i.e., ground truth), in Figure~\ref{fig:bar_stacked_model_accuracy} we grouped participants' responses by \ac{tsv} and reported the simple accuracy of the different \ac{pmv} formulations over each sub-figure.
Both \ac{pmv} formulations underestimated the conditions at which participants are thermally dissatisfied with their environment.
The overall accuracy of the \ac{pmv} and \ac{pmv-ce} were \qty{33}{\percent} and \qty{34}{\percent}, respectively.
Being the classification problem unbalanced, the simple accuracy is not an optimal metric to rate the model prediction accuracy since a default strategy of guessing the majority class would lead to a high overall model accuracy.
A model that would have always predicted `neutral' would have achieved an overall accuracy of \var{perc_tsv_neutral}, the number of participants who voted \ac{tsv}=0.
For this reason, in Table~\ref{tab:f1}, we report the F1-scores for all the models.
The F1-macro score, which is free from label imbalance, depicts that the accuracy of both models across all metrics is marginally better than random guessing (i.e., \qty{14}{\percent}).
% todo:
%  in the table I should also report the F1 scores if I filter out some of the TSV.
%  The PMV model should only be used in the range -2 and 2 so at least we should report also the results in this range.
\begin{table}[htb!]
    \centering
    \input{tables/f1.tex}
    \caption{F1-score for the \ac{pmv} and \ac{pmv-ce} models.}
    \label{tab:f1}
\end{table}

The \ac{pmv-ce} was specifically developed to more accurately estimate latent and sensible heat losses from the skin to the environment when air movement is used~\cite{arens_moving_2009}.
However, our results show that the \ac{pmv-ce} is less accurate than the \ac{pmv} in predicting people who were `slightly warm'.
Both models had an accuracy lower than \qty{1}{\percent} when predicting either `warm', `hot', or `cold'.
This value is significantly lower and worse than random guessing, this is very concerning because this means that the models are actively misleading the users.
The low prediction accuracy of both model formulations, coupled with the lack of data in the extremes of the applicability limits of the models, may suggest that the models' applicability limits should be revisited.
Based on the results, we suggest that the Standards applicability limits could be reduced to the following range \qty{18}{\celsius}~$\leq$~\ac{tdb}~$\leq$~\qty{30}{\celsius} and \qty{10}{\celsius}~$\leq$~\ac{tr}~$\leq$~\qty{30}{\celsius} (instead of \qty{10}{\celsius}~$\leq$~\ac{tdb}~$\leq$~\qty{30}{\celsius} and \qty{10}{\celsius}~$\leq$~\ac{tr}~$\leq$~\qty{40}{\celsius}),  until more data are collected.

@@

\subsection{Comparison between the PMV formulations with the two-node model}\label{subsec:comparison-between-the-pmv-formulations-wih-the-two-node-model}
We compared how the results of the \ac{pmv} and \ac{pmv-ce} perform against the results of the two-node model because xxxx.
Figure~\ref{fig:pmv_two_node_comparison} was generated by calculating the three \ac{pmv} values using the same combination of \num{10000} randomly generated sets inputs.
The \ac{pmv} and \ac{pmv-ce} were then plotted on the y-axis while the \ac{pmvg} value was on the x-axis.
The Figure also shows the \ac{lowess} curves that are used to visualize the prevailing data trend.
The results of the \ac{pmv} model had a higher agreement than those obtained with the \ac{pmv-ce} model for \ac{pmvg} $\geq$ \qty{0.5}{\m\per\s}.
These results show the \ac{pmv-ce} is less similar to the \ac{pmvg} model than the \ac{pmv}, despite using it in the backend to calculate the cooling effect.
This is also an unexpected result and we will discuss why this may be in the following paragraph. 
% todo When a figure report how well a variable is predicted, for example x-measured vs x-predicted, the graph must be a square, otherwise it is visually skewed. 1) I suggest to use the words instead of numbers for PMV 2) We should be consistent with the PMV naming we used (do not use ASHRAE and ISO here) 3) why both models are far from Gagge for PMV<0, both model in general produce a cooler feeling than PMVgagge, we should explain why if we keep this paper.

>>>>

\subsection{\ac{db2} overview}\label{subsec:comfort-db-overview}
While the \ac{db2} contains \var{entries_db_all} data points, only \var{entries_db_used} entries met the inclusion criteria listed in the Methodology section.
The distribution of the six input variables used to calculate the PMV indices is visualized in Figure~\ref{fig:dist_input_data}.
The 2.5th, 50th (median), and 97.5th percentiles are shown in blue.
% TODO:
%  azzurrino or golden or marrone
% remove numbers and stop at 97.5 with the box and 2.5
% max ticks on the y line is 3 to 5, in some figure I have 10
\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/dist_input_data}
    \caption{Distribution of the input variables used to calculate the \ac{pmv} and \ac{pmv-ce} values.
    The text in blue is showing the 2.5th, 50th (median), and 97.5th percentiles.
    The data are shown using boxen plots (letter-value plots).
    They depict the median as the center-line and each successive level outward contains half of the remaining data.}
    \label{fig:dist_input_data}
\end{figure*}
The input data distributions are far from uniformity (equal density) across the Standards' applicability ranges.
We were not able to determine the accuracy of the \ac{pmv} and \ac{pmv-ce} formulations for inputs that were lower than the 2.5th and 97.5th percentiles, since we only had a limited number of data points beyond those values.
For example, only 2.5\% of the available dataset had a \ac{tdb} $\leq$ than \var{ta_95_perc_min}, this may consequently be an issue since both thermal comfort Standards' applicability limits extend to \qty{10}{\celsius}.
This can be explained by the fact that only a few buildings in the world operate at \ac{tdb} lower than this value.
This is because an indoor \ac{tdb} lower than \var{ta_95_perc_min} would be deemed to be unsatisfactory by a great number of occupants~\cite{iso7730}.
This lack of data in the \ac{db2} is particularly relevant for \ac{v}.
The \ac{pmv} and \ac{pmv-ce} only differ when the value of \ac{v} $\geq$ \qty{0.1}{\m\per\s}, however, the median value in the \ac{db2} was \qty{0.08}{\m\per\s} suggesting that most of the data points were collected in buildings with low air movement.
Consequently, we could not reliably test the accuracy of the \ac{pmv} formulations for values of \ac{v} $\geq$ \var{v_95_perc_max}.

Figure~\ref{fig:dist_other_data} shows the distribution of age, height, weight, and running mean outdoor temperature.
Less than half (\num{23300}) of the total entries have information about the participant's sex and are almost equally distributed among males \qty{52}{\percent} and females.
Ages are not normally distributed, the same is true for the running mean outdoor temperature.
About half of the participants (\qty{52}{\percent}) were aged between \num{20} and \num{35} years old, and only \qty{2.5}{\percent} were older than 60.
This limits our analysis to young adults and does not allow us to determine the accuracy of the \ac{pmv} and \ac{pmv-ce} models in predicting thermal sensation for children or older adults.
The \ac{db2} does not contain information about the health status of the participants.
Approximately \qty{56}{\percent} of the running mean outdoor temperature values where between \qtyrange{10}{25}{\celsius}.
% todo:
%  I merged men and women since not considering sex difference in the paper.
%  This figure can be merged with figure 2 because it is a visualization of the database.
%  To be more consisten in the paper, I would use the letter value plots for this and the figure in section 3.3.2
% remove the gridlines or marker them very short;
\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/dist_other_data}
    \caption{Distribution of age, height, weight, and running mean outdoor temperature.
    The data are grouped by sex.
    The text in blue is showing the 2.5th, 50th (median), and 97.5th percentiles.}
    \label{fig:dist_other_data}
\end{figure*}
The percentage of \ac{tpv} grouped by each \ac{tsv} is shown on the left side of Figure~\ref{fig:bar_plot_tp_by_ts}.
Only \var{entries_with_tp} entries had information about both \ac{tsv} and \ac{tpv}.
Thus, on the right, we show a bar chart depicting the distribution of all the \ac{tsv} votes, we analyzed in the paper.
The \ac{tsv} is unbalanced. 
Approximately \var{perc_tsv_neutral} of all the entries have a \ac{tsv} of `Neutral'.
While less than \var{perc_tsv_hot} of the total sample of participants reported to be either `Hot' or `Cold'.
% todo:
% barre piu spesse
% y axis less tick
% use K instead of 000
% remove gridlines
% use Toby's colour

\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/bar_plot_tp_by_ts}
    \caption{The left Figure shows the percentage of \ac{tpv} for each thermal sensation vote.
    The numbers on the right side of each bar show the number of points for each \ac{tsv}.
    The right figure shows the total number of data points grouped by \ac{tsv}.
    Each bin in the right figure has more data points than in the left figure since less data on \ac{tpv} were available.}
    \label{fig:bar_plot_tp_by_ts}
\end{figure*}

@@

\subsection{Comparison of PMV accuracy in predicting thermal sensation}\label{subsec:model-accuracy-comparison-in-predicting-thermal-sensation}
The \ac{pmv} was developed with the primary aim of predicting \ac{tsv}, consequently, we are here comparing the prediction accuracy of the two formulations.
% todo:
%  We can use Toby's 
%  call charts a,b,c,d
%  and have a thin grey line to separate the two group, another option would be to add a light grey background for the two PMV and show that they are distinct.
\begin{figure*}[htb!]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/bar_stacked_model_accuracy_0}
        \caption{All data}
     \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/bar_stacked_model_accuracy_0.2}
        \caption{Only data with \ac{v} $\geq$ \qty{0.2}{\m\per\s}}
     \end{subfigure}
    \caption{The calculated \ac{pmv} value is grouped by the \ac{tsv} value reported by the participants. 
    The number of data points in each bin of the two charts is the same and this allows us to compare the accuracy of both models.
    The overall prediction accuracy is shown above the chart while the model accuracy of predicting participants grouped by their \ac{tsv} is overlayed over each bar. 
    A random model, randomly selecting a thermal sensation vote would have an overall accuracy of \qty{14}{\percent}.}
    \label{fig:bar_stacked_model_accuracy}
\end{figure*}
Being \ac{tsv} the value reported by participants (i.e., ground truth), in Figure~\ref{fig:bar_stacked_model_accuracy} we grouped participants' responses by \ac{tsv} and reported the simple accuracy of the different \ac{pmv} formulations over each sub-figure.
Both \ac{pmv} formulations underestimated the conditions at which participants are thermally dissatisfied with their environment.
The overall accuracy of the \ac{pmv} and \ac{pmv-ce} were \qty{33}{\percent} and \qty{34}{\percent}, respectively.
Being the classification problem unbalanced, the simple accuracy is not an optimal metric to rate the model prediction accuracy since a default strategy of guessing the majority class would lead to a high overall model accuracy.
A model that would have always predicted `neutral' would have achieved an overall accuracy of \var{perc_tsv_neutral}, the number of participants who voted \ac{tsv}=0.
For this reason, in Table~\ref{tab:f1}, we report the F1-scores for all the models.
The F1-macro score, which is free from label imbalance, depicts that the accuracy of both models across all metrics is marginally better than random guessing (i.e., \qty{14}{\percent}).
% todo:
%  in the table I should also report the F1 scores if I filter out some of the TSV.
%  The PMV model should only be used in the range -2 and 2 so at least we should report also the results in this range.
% consider limiting the table to PMV = 0 or PMV -1
\begin{table}[htb!]
    \centering
    \input{tables/f1.tex}
    \caption{F1-score for the \ac{pmv} and \ac{pmv-ce} models.}
    \label{tab:f1}
\end{table}

The \ac{pmv-ce} was specifically developed to more accurately estimate latent and sensible heat losses from the skin to the environment when air movement is used~\cite{arens_moving_2009}.
Both models had an accuracy lower than \qty{1}{\percent} when predicting either `warm', `hot', or `cold'.
This value is significantly lower and worse than random guessing, this is very concerning because this means that the models are actively misleading the users.
The low prediction accuracy of both model formulations, coupled with the lack of data in the extremes of the applicability limits of the models, may suggest that the models' applicability limits should be revisited.
% todo call it Standards implications, limit use PMV and of environmental variables
Based on the results, we suggest that the Standards applicability limits could be reduced to the following range \qty{18}{\celsius}~$\leq$~\ac{tdb}~$\leq$~\qty{30}{\celsius} and \qty{10}{\celsius}~$\leq$~\ac{tr}~$\leq$~\qty{30}{\celsius} (instead of \qty{10}{\celsius}~$\leq$~\ac{tdb}~$\leq$~\qty{30}{\celsius} and \qty{10}{\celsius}~$\leq$~\ac{tr}~$\leq$~\qty{40}{\celsius}),  until more data are collected.

@@

\subsection{Comparison between the PMV formulations with the two-node model}\label{subsec:comparison-between-the-pmv-formulations-wih-the-two-node-model}
% todo consider moving to appendix and explain well why I have done this.
We compared how the results of the \ac{pmv} and \ac{pmv-ce} perform against the results of the two-node model because xxxx.
Figure~\ref{fig:pmv_two_node_comparison} was generated by calculating the three \ac{pmv} values using the same combination of \num{10000} randomly generated sets inputs.
The \ac{pmv} and \ac{pmv-ce} were then plotted on the y-axis while the \ac{pmvg} value was on the x-axis.
The Figure also shows the \ac{lowess} curves that are used to visualize the prevailing data trend.
The results of the \ac{pmv} model had a higher agreement than those obtained with the \ac{pmv-ce} model for \ac{pmvg} $\geq$ \qty{0.5}{\m\per\s}.
These results show the \ac{pmv-ce} is less similar to the \ac{pmvg} model than the \ac{pmv}, despite using it in the backend to calculate the cooling effect.
This is also an unexpected result and we will discuss why this may be in the following paragraph. 
% todo When a figure report how well a variable is predicted, for example x-measured vs x-predicted, the graph must be a square, otherwise it is visually skewed. 1) I suggest to use the words instead of numbers for PMV 2) We should be consistent with the PMV naming we used (do not use ASHRAE and ISO here) 3) why both models are far from Gagge for PMV<0, both model in general produce a cooler feeling than PMVgagge, we should explain why if we keep this paper.
